---
title: "final_project_dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
```

Column {data-width=650}
-----------------------------------------------------------------------

### Chart A

```{r 1}

```



### Chart B

```{r 2}

```

Column {data-width=350}
-----------------------------------------------------------------------
### Chart C

```{r 3}

```

### Chart D
```{r 4 cl3664}
cod_datayr2 = read_csv("./data/NCHS_-_Potentially_Excess_Deaths_from_the_Five_Leading_Causes_of_Death.csv") %>%
  clean_names() %>%
  na.omit() %>%
  filter(!(state == "United States")) %>%
  separate(., percent_potentially_excess_deaths, into = c("percent_excess_death"), sep = "%") %>%
  mutate(percent_excess_death = as.numeric(percent_excess_death), mortality = observed_deaths/population * 10000, mortality = as.numeric(mortality)) %>%
  select(year, cause_of_death, state, benchmark, locality, observed_deaths, population, expected_deaths, potentially_excess_deaths, percent_excess_death, mortality)

lineplotyr = cod_datayr2%>%
  filter(benchmark == "2010 Fixed")%>%
  group_by(cause_of_death,year, locality)%>%
  mutate(year_percent_death=mean(percent_excess_death))%>%
  distinct(year, cause_of_death, locality, .keep_all = TRUE)%>%
  ungroup(year, locality)%>%
  ggplot(aes(x = year, y = year_percent_death, color = cause_of_death))+
  geom_point(size = 0.5)+
  geom_line(size = 0.5) +
  facet_grid(.~locality)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_x_continuous( breaks=seq(2005,2015,1))+
  labs(title = "Average yearly excess death percentage for different locality") 
ggplotly(lineplotyr)
```

As can be seen in the plot, if we stratify by cause, we will find some increasing pattern in subgroups(unintentional injury) in certain period of time(2010-2015). In nonmetropolitan regions, unintentional injury and chronic lower respiratory disease have on average the highest excess death percentage in nonmetropolitan areas and needs more attention from public health experts.