---
title: "p8105_final_project"
author: "Yulan Zhang"
date: "November 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load lib, include=FALSE}
library(tidyverse)
library(rvest)
library(httr)
library(janitor)
library(stringr)
library(readxl)
```


```{r load data}
cod_data = read_csv("./data/NCHS_-_Potentially_Excess_Deaths_from_the_Five_Leading_Causes_of_Death.csv") %>%
  clean_names() %>%
  na.omit() %>%
  filter(!(state == "United States")) %>%
  separate(., percent_potentially_excess_deaths, into = c("percent_excess_death"), sep = "%") %>%
  mutate(percent_excess_death = as.numeric(percent_excess_death), mortality = observed_deaths/population * 10000, mortality = as.numeric(mortality)) %>%
  select(year, cause_of_death, state, locality, observed_deaths, population, expected_deaths, potentially_excess_deaths, percent_excess_death, mortality)

##columns removed

 #"state_fips_code"           "hhs_region"                "age_range"    "benchmark"   "potentially_excess_deaths" "percent_excess_death"      "mortality"   

```
- the `year` variable contains data collected from 2005-2015.
- filtered out `United States` in the `state` variable.
- Added a variable `mortality` which is calculated by observed_deaths/population * 10000. This variable indicates the number of deathes observed in every 10000 people in the three geographic regions: `Metropolitan`, `Nonmetropolitan` and `All`.  



xs2329:
This bar graph shows the distribution of cause of death within mortality in the three geographic regions: `Metropolitan`, `Nonmetropolitan` and `All`. 


```{r xs2329 plot}
cod_data %>%
  group_by(cause_of_death) %>% 
  ggplot(aes(x = locality, y = mortality, fill = cause_of_death)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title="Locality vs. Mortality") +
  labs(x="locality", y="mortality") 
```

```{r xs2329 boxplot}
cod_data %>%
  group_by(cause_of_death) %>% 
  ggplot(aes(x = locality, y = mortality)) + geom_boxplot(aes(color = cause_of_death), na.rm = T) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title="Locality vs. Mortality") +
  labs(x="locality", y="mortality") 
```



```{r xs2329 lm}
mortality_lm = lm(mortality ~locality, data = cod_data)
summary(mortality_lm)
```
interpretation on linear regression:
We expect to see 0.22576 less deaths for every 10000 people in metropolitan region as compared to all regions. 
We expect to see 1.14451 more death for every 10000 people in nonmetropolitan region as compared to all regions.

<<<<<<< HEAD
```{r wp2241}
### Question of concern: If percent excess death has significant difference between metro and non-metro groups.
gp_cod_data = cod_data %>%
  mutate(percent_excess_death = as.numeric(percent_excess_death)) %>%
  group_by(locality, year)%>% 
  mutate(mean = mean(percent_excess_death)) #doesn't work. Ask TA.

ggplot(gp_cod_data, aes(x = year, y = mean, fill = locality)) + geom_bar(stat = "identity")+
facet_grid(. ~ locality) 



```


=======


<<<<<<< HEAD
=======
<<<<<<< HEAD
```{r cl3664 plot}
cod_datayr = cod_data
cod_datayr%>%
  filter(!is.na(observed_deaths),
         state != "United States")%>%

  ggplot(aes(x = year, y = percent_excess_death, group = cause_of_death))+
geom_bar(aes(fill = cause_of_death), stat = "identity")
```

```{r cl3664 lm}

```


```{r yz3306.1}
#For each of the five cause of death, what is the mean percent of excess death? It is different across locality.
library(forcats)
names(cod_data)
cod_data %>%
  na.omit %>%
  group_by(cause_of_death) %>%
  mutate(mean_percent_excess_death = mean(percent_excess_death)) %>%
  ungroup(cause_of_death) %>%
  mutate(cause_of_death = fct_reorder(cause_of_death, mean_percent_excess_death)) %>%
  ggplot(aes(x = cause_of_death, y = percent_excess_death, fill = cause_of_death)) +
  geom_boxplot() +
  facet_grid(~locality) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))  +
  theme(legend.position = "bottom")

##As shown in the boxplots for all three regions, the rank of the five cause of death remain the same for different localities. For all five causes, the mean percent of death is lower for `Metropolitan` and higher for `Nonmetropolotan`, with respect to `All` Locality.
```

```{r yz3306.2}
#regress percent of excess death on cause of death, which is a categorical variable.
#with respect to cancer, the least of the five causes, all 4 have additional percent of death as evaluated by the estimates. The results give statisically significant estimates 

cod_lm_yz1 = lm(percent_excess_death ~cause_of_death, data = cod_data)
broom::tidy(cod_lm_yz1)

#cod_lm_yz2 = lm(percent_excess_death ~locality, data = cod_data)
#broom::tidy(cod_lm_yz2)

cod_lm_yz3 = lm(percent_excess_death ~cause_of_death + locality, data = cod_data)
broom::tidy(cod_lm_yz3)
```

